---
title: "Annexation Reach and Metro Growth, 2000-2020"
subtitle: "Does a city's share of its home county's population predict its growth rate, controlling for state-line fragmentation and region?"
format:
  html:
    code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ggiraph)
library(modelsummary)
library(kableExtra)
library(RColorBrewer)
library(scales)

theme_set(theme_minimal(base_size = 12))
```

## Data

```{r}
#| label: load-data
#| message: false
#| warning: false

city <- read_csv("data/city_pop_2000_2020.csv", show_col_types = FALSE)
county <- read_csv("data/county_pop_2000_2020.csv", show_col_types = FALSE)
cbsa_flags <- read_csv("data/cbsa_state_flags.csv", show_col_types = FALSE)
```

### Region / division lookup

Standard Census Bureau 4-region grouping, derived from `state_fips`.

```{r}
#| label: region-lookup

region_lookup <- tribble(
  ~state_fips, ~region,
  "09", "Northeast", "23", "Northeast", "25", "Northeast", "33", "Northeast",
  "44", "Northeast", "50", "Northeast", "34", "Northeast", "36", "Northeast",
  "42", "Northeast",
  "18", "Midwest", "17", "Midwest", "26", "Midwest", "39", "Midwest",
  "55", "Midwest", "19", "Midwest", "20", "Midwest", "27", "Midwest",
  "29", "Midwest", "31", "Midwest", "38", "Midwest", "46", "Midwest",
  "10", "South", "11", "South", "12", "South", "13", "South", "24", "South",
  "37", "South", "45", "South", "51", "South", "54", "South", "01", "South",
  "21", "South", "28", "South", "47", "South", "05", "South", "22", "South",
  "40", "South", "48", "South",
  "04", "West", "08", "West", "16", "West", "35", "West", "30", "West",
  "49", "West", "32", "West", "56", "West", "02", "West", "06", "West",
  "15", "West", "41", "West", "53", "West",
  "72", "Puerto Rico"
)
```

### Build the analysis panel

Join city -> county (by `county_fips`) and city -> CBSA state flags (by `cbsa_code`), compute `pct_of_county` and `cagr`.

```{r}
#| label: build-panel

panel <- city %>%
  left_join(
    county %>% select(county_fips, county_name, population_2020_county = population_2020),
    by = "county_fips"
  ) %>%
  left_join(
    cbsa_flags %>% select(cbsa_code, n_states, multi_state_metro),
    by = "cbsa_code"
  ) %>%
  left_join(region_lookup, by = "state_fips") %>%
  mutate(
    pct_of_county = population_2020 / population_2020_county * 100,
    cagr = (population_2020 / population_2000)^(1 / 20) - 1,
    multi_state_metro_flag = as.integer(multi_state_metro),
    region = factor(region, levels = c("South", "Northeast", "Midwest", "West", "Puerto Rico"))
  )

# regression sample: drop the 6 rows with no usable 2000 population
panel_reg <- panel %>% filter(!is.na(population_2000))

cat("Full panel:", nrow(panel), "rows\n")
cat("Regression sample (population_2000 non-missing):", nrow(panel_reg), "rows\n")
cat("Excluded (no 2000 population):", nrow(panel) - nrow(panel_reg), "rows ->",
    paste(panel %>% filter(is.na(population_2000)) %>% pull(place_name), collapse = ", "))
```

## Regression

`cagr ~ pct_of_county + multi_state_metro + region`, with metro total population (log) added as a covariate in a second specification.

```{r}
#| label: model-fit

m1 <- lm(cagr ~ pct_of_county + multi_state_metro_flag, data = panel_reg)
m2 <- lm(cagr ~ pct_of_county + multi_state_metro_flag + region, data = panel_reg)

panel_reg <- panel_reg %>%
  mutate(log_metro_pop = log(population_2020_county))

m3 <- lm(cagr ~ pct_of_county + multi_state_metro_flag + region + log_metro_pop, data = panel_reg)

modelsummary(
  list("Baseline" = m1, "+ Region" = m2, "+ Region + Metro Pop (log)" = m3),
  stars = TRUE,
  coef_rename = c(
    pct_of_county = "City % of county population",
    multi_state_metro_flag = "Multi-state metro (0/1)",
    regionNortheast = "Region: Northeast",
    regionMidwest = "Region: Midwest",
    regionWest = "Region: West",
    log_metro_pop = "log(county population, 2020)"
  ),
  gof_omit = "AIC|BIC|Log.Lik|F",
  output = "kableExtra"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Does the annexation-reach effect on `pct_of_county` survive controlling for region (and metro size)?** Yes on significance, but the sign is the opposite of the naive "bigger annexation reach -> faster growth" story: `pct_of_county` is negative and significant (p < 0.05) in all three specifications, meaning cities that already hold a *larger* share of their county's population tend to have *slower* 2000-2020 CAGR, controlling for region and metro size — consistent with a ceiling/mean-reversion effect (there is less room left to annex or in-fill once a city already dominates its county) rather than annexation reach driving extra growth. `multi_state_metro_flag` is never statistically distinguishable from zero and even flips sign once region is added, so on this cross-section state-line fragmentation alone does not show up as an independent growth penalty net of region and `pct_of_county` — the Columbus/Cincinnati contrast below should be read as a single illustrative case, not as evidence of a general multi-state effect.

## Growth vs. annexation reach (interactive)

```{r}
#| label: scatter-plot
#| fig-width: 8
#| fig-height: 6

panel_reg <- panel_reg %>%
  mutate(
    multi_state_lab = if_else(multi_state_metro, "Multi-state metro", "Single-state metro"),
    tooltip_text = paste0(
      place_name, ", ", state_name, "\n",
      "% of county pop (2020): ", number(pct_of_county, accuracy = 0.1), "%\n",
      "CAGR 2000-2020: ", percent(cagr, accuracy = 0.01), "\n",
      multi_state_lab
    )
  )

p <- ggplot(panel_reg, aes(x = pct_of_county, y = cagr, color = multi_state_lab)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text, data_id = place_geoid),
    size = 2.2, alpha = 0.8
  ) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, color = "grey30", linewidth = 0.6) +
  scale_color_brewer(palette = "Set2", name = NULL) +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    x = "City population as % of home county population (2020)",
    y = "City population CAGR, 2000-2020",
    title = "City growth rate vs. share of home county population"
  ) +
  theme(legend.position = "bottom")

girafe(
  ggobj = p,
  width_svg = 8, height_svg = 6,
  options = list(
    opts_sizing(rescale = TRUE),
    opts_tooltip(opacity = 0.9)
  )
)
```

## Columbus, OH vs. Cincinnati, OH

```{r}
#| label: columbus-cincinnati

cc <- panel %>%
  filter(place_name %in% c("Columbus", "Cincinnati"), state_name == "Ohio") %>%
  select(place_name, state_name, county_name, population_2000, population_2020,
         pct_of_county, multi_state_metro, cagr) %>%
  arrange(place_name)

cc %>%
  mutate(
    population_2000 = comma(population_2000),
    population_2020 = comma(population_2020),
    pct_of_county = number(pct_of_county, accuracy = 0.1),
    cagr = percent(cagr, accuracy = 0.01)
  ) %>%
  rename(
    City = place_name, State = state_name, County = county_name,
    `Pop. 2000` = population_2000, `Pop. 2020` = population_2020,
    `% of county pop.` = pct_of_county, `Multi-state metro` = multi_state_metro,
    `CAGR 2000-2020` = cagr
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

::: {.callout-note}
Motivating case: Columbus (single-state metro, ~68% of Franklin County) vs. Cincinnati (multi-state OH-KY-IN metro, ~37% of Hamilton County). See table above for the paired `pct_of_county` / `multi_state_metro` / `cagr` values.
:::

## Full panel table

```{r}
#| label: full-table

panel %>%
  select(place_name, state_name, cbsa_title, pct_of_county, multi_state_metro,
         population_2000, population_2020, cagr) %>%
  arrange(desc(pct_of_county)) %>%
  mutate(
    population_2000 = comma(population_2000),
    population_2020 = comma(population_2020),
    pct_of_county = number(pct_of_county, accuracy = 0.1),
    cagr = if_else(is.na(cagr), NA_character_, percent(cagr, accuracy = 0.01))
  ) %>%
  rename(
    City = place_name, State = state_name, CBSA = cbsa_title,
    `% of county pop.` = pct_of_county, `Multi-state metro` = multi_state_metro,
    `Pop. 2000` = population_2000, `Pop. 2020` = population_2020,
    `CAGR 2000-2020` = cagr
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  scroll_box(height = "500px")
```

## Discussion

### Annexation reach as positional competition

### State lines as a coordination barrier

### Where this stops short of a full Girardian treatment
