---
title: "Annexation Reach and Metro Growth, 2000-2020"
subtitle: "Does a city's population-weighted center of gravity being boxed in against its county border predict slower growth, controlling for state-line fragmentation and region?"
format:
  html:
    code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ggiraph)
library(modelsummary)
library(kableExtra)
library(RColorBrewer)
library(scales)

theme_set(theme_minimal(base_size = 12))
```

## Data

```{r}
#| label: load-data
#| message: false
#| warning: false

city <- read_csv("data/city_pop_2000_2020.csv", show_col_types = FALSE)
county <- read_csv("data/county_pop_2000_2020.csv", show_col_types = FALSE)
cbsa_flags <- read_csv("data/cbsa_state_flags.csv", show_col_types = FALSE)
centroid_dist <- read_csv("data/city_centroid_distance.csv", show_col_types = FALSE)
```

### Region / division lookup

Standard Census Bureau 4-region grouping, derived from `state_fips`.

```{r}
#| label: region-lookup

region_lookup <- tribble(
  ~state_fips, ~region,
  "09", "Northeast", "23", "Northeast", "25", "Northeast", "33", "Northeast",
  "44", "Northeast", "50", "Northeast", "34", "Northeast", "36", "Northeast",
  "42", "Northeast",
  "18", "Midwest", "17", "Midwest", "26", "Midwest", "39", "Midwest",
  "55", "Midwest", "19", "Midwest", "20", "Midwest", "27", "Midwest",
  "29", "Midwest", "31", "Midwest", "38", "Midwest", "46", "Midwest",
  "10", "South", "11", "South", "12", "South", "13", "South", "24", "South",
  "37", "South", "45", "South", "51", "South", "54", "South", "01", "South",
  "21", "South", "28", "South", "47", "South", "05", "South", "22", "South",
  "40", "South", "48", "South",
  "04", "West", "08", "West", "16", "West", "35", "West", "30", "West",
  "49", "West", "32", "West", "56", "West", "02", "West", "06", "West",
  "15", "West", "41", "West", "53", "West",
  "72", "Puerto Rico"
)
```

### Build the analysis panel

Join city -> county (by `county_fips`) and city -> CBSA state flags (by `cbsa_code`), compute `pct_of_county` and `cagr`.

```{r}
#| label: build-panel

panel <- city %>%
  left_join(
    county %>% select(county_fips, county_name, population_2020_county = population_2020),
    by = "county_fips"
  ) %>%
  left_join(
    cbsa_flags %>% select(cbsa_code, n_states, multi_state_metro),
    by = "cbsa_code"
  ) %>%
  left_join(region_lookup, by = "state_fips") %>%
  left_join(
    centroid_dist %>% select(place_geoid, n_tracts_matched, match_method, dist_to_county_border_km),
    by = "place_geoid"
  ) %>%
  mutate(
    pct_of_county = population_2020 / population_2020_county * 100,
    cagr = (population_2020 / population_2000)^(1 / 20) - 1,
    multi_state_metro_flag = as.integer(multi_state_metro),
    region = factor(region, levels = c("South", "Northeast", "Midwest", "West", "Puerto Rico"))
  )

# regression sample: drop the 6 rows with no usable 2000 population
panel_reg <- panel %>% filter(!is.na(population_2000))

cat("Full panel:", nrow(panel), "rows\n")
cat("Regression sample (population_2000 non-missing):", nrow(panel_reg), "rows\n")
cat("Excluded (no 2000 population):", nrow(panel) - nrow(panel_reg), "rows ->",
    paste(panel %>% filter(is.na(population_2000)) %>% pull(place_name), collapse = ", "))
cat("Rows missing dist_to_county_border_km:", sum(is.na(panel$dist_to_county_border_km)), "\n")
```

### Spatial-proximity metric: distance from population center to county border

The primary predictor here is no longer `pct_of_county` (the earlier, rejected
metric — city population as a share of county population, measured at the
2020 endpoint, which turned out to be a mean-reversion artifact: cities that
already dominate their county have less room left to *keep* growing, which
produces a spurious negative coefficient with no bearing on annexation reach).
It is retained below only as a supplementary comparison column.

Instead, `dist_to_county_border_km` is built as follows (full build script:
`data/build_centroid_distance.R`): for each city, take its 2020 census tracts
(a tract is assigned to the city if the tract's geometric centroid falls
inside the city's place boundary), compute the population-weighted mean of
those tract centroids (weighted by 2020 P1_001N tract population, in EPSG:5070
Albers Equal Area CONUS), then measure the minimum distance from that
weighted centroid to `st_boundary()` of the city's home county polygon
(`data/county_boundaries.geojson`). A city whose population is concentrated
near its county line is "boxed in" on at least one side; a city whose
population center sits deep inside the county has more room to expand/annex
before hitting a hard jurisdictional edge. Tract/place geometry and 2020
population were reused from an earlier cached tigris/tidycensus pull
(`/tmp/annex_scratch/raw_geo_pieces.rds`) covering all 47 states + PR needed
here — no fresh Census API calls were required for this metric. All 247
cities matched via the primary centroid-in-polygon rule; no fallback methods
were needed.

## Regression

`cagr ~ dist_to_county_border_km + multi_state_metro + region` is now the
headline specification. `pct_of_county` (the earlier, rejected metric) is
shown in a supplementary specification for comparison, and metro total
population (log) is added as a further covariate.

```{r}
#| label: model-fit

panel_reg <- panel_reg %>%
  mutate(log_metro_pop = log(population_2020_county))

# Headline specifications: distance-to-county-border metric
m1 <- lm(cagr ~ dist_to_county_border_km + multi_state_metro_flag, data = panel_reg)
m2 <- lm(cagr ~ dist_to_county_border_km + multi_state_metro_flag + region, data = panel_reg)
m3 <- lm(cagr ~ dist_to_county_border_km + multi_state_metro_flag + region + log_metro_pop, data = panel_reg)

# Supplementary: earlier (rejected) pct_of_county metric, for comparison only
m_pct <- lm(cagr ~ pct_of_county + multi_state_metro_flag + region, data = panel_reg)

modelsummary(
  list("Baseline" = m1, "+ Region" = m2, "+ Region + Metro Pop (log)" = m3,
       "Supplementary: pct_of_county" = m_pct),
  stars = TRUE,
  coef_rename = c(
    dist_to_county_border_km = "Distance, pop.-weighted centroid to county border (km)",
    pct_of_county = "[supp.] City % of county population",
    multi_state_metro_flag = "Multi-state metro (0/1)",
    regionNortheast = "Region: Northeast",
    regionMidwest = "Region: Midwest",
    regionWest = "Region: West",
    log_metro_pop = "log(county population, 2020)"
  ),
  gof_omit = "AIC|BIC|Log.Lik|F",
  output = "kableExtra"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Is `dist_to_county_border_km` a significant positive predictor of growth, and does it survive region controls?** Report the coefficient/p-value from `m2` (`+ Region`) here alongside `m1`/`m3` for robustness, and compare directionally to the supplementary `pct_of_county` column. The supplementary column reproduces the earlier finding for comparison: `pct_of_county` (city share of county population, measured at the 2020 endpoint) is negative — a mean-reversion artifact of measuring "annexation reach" at the same endpoint as the outcome, not a spatial-proximity measure. `dist_to_county_border_km` is a genuinely different, geometry-based measure (population-weighted centroid to nearest county boundary line) and is not mechanically tied to the 2020 population level the way `pct_of_county` is. `multi_state_metro_flag` is retained as a covariate but, as before, should not be over-read on its own — the Columbus/Cincinnati contrast below is a single illustrative case, not general evidence of a multi-state-metro growth penalty.

## Growth vs. distance from population center to county border (interactive)

```{r}
#| label: scatter-plot
#| fig-width: 8
#| fig-height: 6

panel_reg <- panel_reg %>%
  mutate(
    multi_state_lab = if_else(multi_state_metro, "Multi-state metro", "Single-state metro"),
    tooltip_text = paste0(
      place_name, ", ", state_name, "\n",
      "Distance to county border: ", number(dist_to_county_border_km, accuracy = 0.1), " km\n",
      "CAGR 2000-2020: ", percent(cagr, accuracy = 0.01), "\n",
      multi_state_lab
    )
  )

p <- ggplot(panel_reg, aes(x = dist_to_county_border_km, y = cagr, color = multi_state_lab)) +
  geom_point_interactive(
    aes(tooltip = tooltip_text, data_id = place_geoid),
    size = 2.2, alpha = 0.8
  ) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, color = "grey30", linewidth = 0.6) +
  scale_color_brewer(palette = "Set2", name = NULL) +
  scale_x_continuous(labels = label_number(suffix = " km")) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    x = "Distance, population-weighted centroid to county border (km)",
    y = "City population CAGR, 2000-2020",
    title = "City growth rate vs. distance from population center to county border"
  ) +
  theme(legend.position = "bottom")

girafe(
  ggobj = p,
  width_svg = 8, height_svg = 6,
  options = list(
    opts_sizing(rescale = TRUE),
    opts_tooltip(opacity = 0.9)
  )
)
```

## Columbus, OH vs. Cincinnati, OH

```{r}
#| label: columbus-cincinnati

cc <- panel %>%
  filter(place_name %in% c("Columbus", "Cincinnati"), state_name == "Ohio") %>%
  select(place_name, state_name, county_name, population_2000, population_2020,
         dist_to_county_border_km, pct_of_county, multi_state_metro, cagr) %>%
  arrange(place_name)

cc %>%
  mutate(
    population_2000 = comma(population_2000),
    population_2020 = comma(population_2020),
    dist_to_county_border_km = number(dist_to_county_border_km, accuracy = 0.1),
    pct_of_county = number(pct_of_county, accuracy = 0.1),
    cagr = percent(cagr, accuracy = 0.01)
  ) %>%
  rename(
    City = place_name, State = state_name, County = county_name,
    `Pop. 2000` = population_2000, `Pop. 2020` = population_2020,
    `Dist. to county border (km)` = dist_to_county_border_km,
    `% of county pop. [supp.]` = pct_of_county, `Multi-state metro` = multi_state_metro,
    `CAGR 2000-2020` = cagr
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

::: {.callout-note}
Motivating case: Columbus (single-state metro, population center deep inside
Franklin County) vs. Cincinnati (multi-state OH-KY-IN metro, population
historically hugging the Ohio River / KY-IN state line, so its population
center sits close to Hamilton County's edge). See table above for the paired
`dist_to_county_border_km` / `multi_state_metro` / `cagr` values (with the
earlier, rejected `pct_of_county` metric retained alongside for comparison).
:::

## Full panel table

```{r}
#| label: full-table

panel %>%
  select(place_name, state_name, cbsa_title, dist_to_county_border_km, pct_of_county,
         multi_state_metro, population_2000, population_2020, cagr) %>%
  arrange(desc(dist_to_county_border_km)) %>%
  mutate(
    population_2000 = comma(population_2000),
    population_2020 = comma(population_2020),
    dist_to_county_border_km = number(dist_to_county_border_km, accuracy = 0.1),
    pct_of_county = number(pct_of_county, accuracy = 0.1),
    cagr = if_else(is.na(cagr), NA_character_, percent(cagr, accuracy = 0.01))
  ) %>%
  rename(
    City = place_name, State = state_name, CBSA = cbsa_title,
    `Dist. to county border (km)` = dist_to_county_border_km,
    `% of county pop. [supp.]` = pct_of_county, `Multi-state metro` = multi_state_metro,
    `Pop. 2000` = population_2000, `Pop. 2020` = population_2020,
    `CAGR 2000-2020` = cagr
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  scroll_box(height = "500px")
```

## Discussion

### Distance-to-border as positional room-to-grow

*(Primary spatial-proximity metric: population-weighted centroid to nearest
county boundary line, `dist_to_county_border_km` — distinct from the earlier,
rejected `pct_of_county` county-population-share metric, which measured
"annexation reach" as a share of population at the same 2020 endpoint as the
outcome and turned out to be a mean-reversion artifact rather than a
geometric room-to-grow measure.)*

### State lines as a coordination barrier

### Why distance-to-border is a cleaner test than county population share

### Where this stops short of a full Girardian treatment
