---
title: "Mapping Annexation Reach: Population Centers, City Boundaries, County Lines"
subtitle: "Top 100 Cities by 2020 Population, Plus Columbus and Cincinnati, OH"
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)
library(RColorBrewer)
library(scales)

sf_use_s2(TRUE)
```

## Data

Same panel as `analysis.qmd` (`city_pop_2000_2020.csv`, `county_pop_2000_2020.csv`,
`cbsa_state_flags.csv`, `city_centroid_distance.csv`), plus two geometry layers:
`data/county_boundaries.geojson` (home-county polygons, already EPSG:4326) and
`data/city_place_boundaries.geojson` (place polygons for the plotted cities,
built by `data/build_place_boundaries.R` by reusing the tigris place cache at
`/tmp/annex_scratch/raw_geo_pieces.rds` rather than a fresh `tigris::places()`
pull -- the same cache `build_centroid_distance.R` used).

```{r}
#| label: load-data
#| message: false
#| warning: false

city <- read_csv("data/city_pop_2000_2020.csv", show_col_types = FALSE)
county <- read_csv("data/county_pop_2000_2020.csv", show_col_types = FALSE)
cbsa_flags <- read_csv("data/cbsa_state_flags.csv", show_col_types = FALSE)
centroid_dist <- read_csv("data/city_centroid_distance.csv", show_col_types = FALSE)

county_bd <- st_read("data/county_boundaries.geojson", quiet = TRUE)
place_bd <- st_read("data/city_place_boundaries.geojson", quiet = TRUE)
```

### Build the same analysis panel as `analysis.qmd`

```{r}
#| label: build-panel

region_lookup <- tribble(
  ~state_fips, ~region,
  "09", "Northeast", "23", "Northeast", "25", "Northeast", "33", "Northeast",
  "44", "Northeast", "50", "Northeast", "34", "Northeast", "36", "Northeast",
  "42", "Northeast",
  "18", "Midwest", "17", "Midwest", "26", "Midwest", "39", "Midwest",
  "55", "Midwest", "19", "Midwest", "20", "Midwest", "27", "Midwest",
  "29", "Midwest", "31", "Midwest", "38", "Midwest", "46", "Midwest",
  "10", "South", "11", "South", "12", "South", "13", "South", "24", "South",
  "37", "South", "45", "South", "51", "South", "54", "South", "01", "South",
  "21", "South", "28", "South", "47", "South", "05", "South", "22", "South",
  "40", "South", "48", "South",
  "04", "West", "08", "West", "16", "West", "35", "West", "30", "West",
  "49", "West", "32", "West", "56", "West", "02", "West", "06", "West",
  "15", "West", "41", "West", "53", "West",
  "72", "Puerto Rico"
)

panel <- city %>%
  left_join(
    county %>% select(county_fips, county_name, population_2020_county = population_2020),
    by = "county_fips"
  ) %>%
  left_join(
    cbsa_flags %>% select(cbsa_code, n_states, multi_state_metro),
    by = "cbsa_code"
  ) %>%
  left_join(region_lookup, by = "state_fips") %>%
  left_join(
    centroid_dist %>% select(place_geoid, n_tracts_matched, match_method,
                              centroid_x_5070, centroid_y_5070, dist_to_county_border_km),
    by = "place_geoid"
  ) %>%
  mutate(
    cagr = (population_2020 / population_2000)^(1 / 20) - 1
  )
```

### Select cities to plot: top 100 by 2020 population, plus Columbus/Cincinnati OH

```{r}
#| label: select-cities

top100_geoids <- panel %>%
  arrange(desc(population_2020)) %>%
  slice_head(n = 100) %>%
  pull(place_geoid)

anchor_geoids <- panel %>%
  filter(place_name %in% c("Columbus", "Cincinnati"), state_name == "Ohio") %>%
  pull(place_geoid)

selected_geoids <- union(top100_geoids, anchor_geoids)

map_panel <- panel %>%
  filter(place_geoid %in% selected_geoids, !is.na(centroid_x_5070), !is.na(centroid_y_5070))

cat("Cities plotted:", nrow(map_panel), "\n")
cat("Home counties among plotted cities:", n_distinct(map_panel$county_fips), "\n")
cat("Place boundaries available:", nrow(place_bd), "\n")
```

### Reproject centroids to WGS84 for leaflet

```{r}
#| label: reproject-centroids

centroid_pts <- map_panel %>%
  filter(!is.na(centroid_x_5070), !is.na(centroid_y_5070)) %>%
  st_as_sf(coords = c("centroid_x_5070", "centroid_y_5070"), crs = 5070) %>%
  st_transform(4326)

coords_ll <- st_coordinates(centroid_pts)
centroid_pts <- centroid_pts %>%
  mutate(lon = coords_ll[, "X"], lat = coords_ll[, "Y"])
```

### Filter county + place boundary layers to the plotted cities' home counties

```{r}
#| label: filter-boundaries

county_bd_sel <- county_bd %>%
  filter(county_fips %in% map_panel$county_fips)

place_bd_sel <- place_bd %>%
  filter(place_geoid %in% map_panel$place_geoid)

cat("County polygons on map:", nrow(county_bd_sel), "\n")
cat("Place polygons on map:", nrow(place_bd_sel), "\n")
```

## Interactive map

Marker color and size are both driven by `dist_to_county_border_km` (the
paper's primary spatial-proximity metric): cities whose population-weighted
center sits close to the county line are "boxed in," cities whose center sits
deep inside the county have more room to expand. County boundaries are drawn
as thin grey outlines with light fill; city place boundaries are drawn as
thicker blue outlines with no fill, so city-within-county containment is
visible at a glance. Columbus and Cincinnati, OH -- the paper's motivating
single-state-vs-multi-state-metro contrast -- get a distinct gold marker and
a bold callout in their popups.

```{r}
#| label: leaflet-map
#| fig-width: 9
#| fig-height: 7

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = centroid_pts$dist_to_county_border_km
)

get_radius <- function(d) {
  scales::rescale(d, to = c(5, 16), from = range(centroid_pts$dist_to_county_border_km, na.rm = TRUE))
}

is_anchor <- centroid_pts$place_geoid %in% anchor_geoids

popup_html <- function(name, state, county, dist_km, cagr, multi_state, anchor) {
  header <- if (anchor) {
    paste0("<b style='font-size:1.05em;color:#b8860b;'>&#9733; ", name, ", ", state, "</b>")
  } else {
    paste0("<b>", name, ", ", state, "</b>")
  }
  HTML(paste0(
    header, "<br/>",
    "Home county: ", county, "<br/>",
    "Distance to county border: ", number(dist_km, accuracy = 0.1), " km<br/>",
    "CAGR 2000-2020: ", percent(cagr, accuracy = 0.01), "<br/>",
    "Multi-state metro: ", if_else(multi_state, "Yes", "No")
  ))
}

popups <- pmap_chr(
  list(centroid_pts$place_name, centroid_pts$state_name, centroid_pts$county_name,
       centroid_pts$dist_to_county_border_km, centroid_pts$cagr,
       centroid_pts$multi_state_metro, is_anchor),
  function(name, state, county, dist_km, cagr, multi_state, anchor) {
    as.character(popup_html(name, state, county, dist_km, cagr, multi_state, anchor))
  }
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = county_bd_sel,
    color = "grey40", weight = 1, opacity = 0.7,
    fillColor = "grey80", fillOpacity = 0.12,
    label = ~county_namelsad,
    group = "County boundaries"
  ) %>%
  addPolygons(
    data = place_bd_sel,
    color = "#2166AC", weight = 2, opacity = 0.9,
    fill = FALSE,
    label = ~place_name_geo,
    group = "City boundaries"
  ) %>%
  addCircleMarkers(
    data = centroid_pts[!is_anchor, ],
    lng = ~lon, lat = ~lat,
    radius = ~get_radius(dist_to_county_border_km),
    color = "black", weight = 1,
    fillColor = ~pal(dist_to_county_border_km),
    fillOpacity = 0.85,
    popup = popups[!is_anchor],
    group = "Population-weighted centroids"
  ) %>%
  addCircleMarkers(
    data = centroid_pts[is_anchor, ],
    lng = ~lon, lat = ~lat,
    radius = ~get_radius(dist_to_county_border_km) + 3,
    color = "black", weight = 2,
    fillColor = "#FFD700",
    fillOpacity = 0.95,
    popup = popups[is_anchor],
    group = "Population-weighted centroids"
  ) %>%
  addLegend(
    "bottomright", pal = pal, values = centroid_pts$dist_to_county_border_km,
    title = "Distance, centroid to<br/>county border (km)",
    opacity = 1
  ) %>%
  addLayersControl(
    overlayGroups = c("County boundaries", "City boundaries", "Population-weighted centroids"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  setView(lng = -93, lat = 38, zoom = 4)
```

::: {.callout-note}
Gold markers are Columbus and Cincinnati, OH. Columbus's population-weighted
centroid sits deep inside Franklin County (single-state metro); Cincinnati's
sits close to Hamilton County's edge, consistent with its historical
orientation toward the Ohio River / KY-IN state line (multi-state metro). See
`analysis.qmd` for the paired regression table and full-panel comparison.
:::
